Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea

### 2022-10-08 analysis/paper/paper.qmd
add map data
fix some bad in text citations

The stemmed point is the first composite tool in the Korean Paleolithic that requires two different parts, consisting of a stone point and shaft, presumably made out of wood [@seong2008tanged]. Using blades enables mass production for this composite tool and its shape becomes more standardized [@lee2015characteristics; @park2022did]. Therefore, -@o2018origins claims that understanding the emergence of projectile technology gives insights to greater cultural, evolutionary, and behavioral cognitive flexibility.

Since the stemmed points represent the beginning of the Korean Late Paleolithic, investigating their origin(s) becomes a key for understanding the technological transition from the Early Paleoltihic, modern human dispersals, and existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern]. There has been a northworthy debate about explaining the origin(s) of stemmed points and it can be summarized into two competing models: 'heterogenic' migration [@bae2010origin], in situ evolution[@seong2009emergence]. The migration model claims that the new blade industry including stemmed points, and the earlier coarse flake tradition including large cores, polyhedrons, choppers, and even handaxes, came from different origins (e.g. 1: North route: Siberia, Mongolia, or other regions of northeast China, and 2: South route: southern China) as the result of continuing influx of modern human migration from two routes. The 'in situ' model argues that stemmed points and other Late Paleolithic technologies, including blade industries, autonomously emerged in the south of the Korean peninsula, with no apparent external influence. The difference between the two models comes from how to understand a few early sites that contain stemmed points but no blades and blade coares. The 'in situ' model claims that the stemmed points appeared before blades by acknowledging those early sites including Bonggok, Songamri, Yonghodong, and Hwadaeri sites while migration model supports the traditional definition of the Late or Upper Palaeolithic in Europe and Africa that started with emergence and establishment of blade technology [@seong2016eastern].

Stemmed points are important proxies to understand human behaviors during the Late Paleolithic. -@park2022did examine mobility and site occupation patterns through lithic assemblages and concepts of human behavioral ecology and find that forager groups using stemmed points may have been associated with occupation of marginal or extreme environments compared to the groups with no stemmed points. Also, they were more associated with expedient technologies, indicating residential and less mobile behaviors. -@Chong_2021 claims that the morphological variation of the stemmed points along with tool types in lithic assemblages, assemblage size, use of raw materials, and types of blanks could represent the characteristics of occupations such as "limited activity station" and "residential base camp." For example, stemmed points with high morphological variations in a tool size, shape of edge, degrees of damage, and types of edge retouching from Yongsandong site indicate that the site was used for specific or limited activities such as hunting [@kim2004yongsan; @seong2015diversity; @Chong_2021].

@@

Though the stemmed points are generally assumed as hunting armatures [@lee2019were; @lee2011study; @seong2008tanged; @lee2002new; @chang2013human], it is difficult to understand the function of the tools without knowing their complete shape attached to other components. Preserved wooden components of projectile tools are too few and rare to generalize the overall scale and variability of the tools used in prehistory [@shea2006origins]. -@Lee_Kong_2002 claimed to name the stemmed points as 'stemmed tools' because of the diverse shape of stemmed points, which could represent different functions as well as the uncertainty of complete tool shape. Other researchers agree that some stemmed points that are not symmetrical and have retouch on one side of the artifacts or have denticulate blades on one side should not be called as stemmed points but stemmed knives, stemmed side-scrapers, stemmed end-scrapers, or stemmed burins [@seong2008tanged; @Lee_Jang_2011; @Kim] @fig-sp-hand-made: 7-8.

@@

The tip cross-sectional area (TCSA) has been used as a ballistically significant standard to discriminate different functional classes of projectile armatures such as spearthrower (a.k.a. atlatl) dart tips, arrowheads and large experimental stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins]. TCSA metric indicates the part of the tool that cuts the target's hide and relates to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation]. This method was first proposed by -@hughes1998getting and and tested by -@shea2006origins through comparing archeological examples to ethnographically collected samples of known use. -@shea2009impact also applied this approach to compare projectile points from Africa, the Levant, and Europe and claimed that projectile weapons first appeared in Africa. The main advantage of the TCSA is, regardless of point type, only maximum width and thickness measurements are needed for calculating the TCSA value (0.5 × maximum width × maximum thickness) [@sisk2011african; @lombard2020tip]. Later, -@sisk2011african proposed an alternate measure, tip cross-sectional perimeter (TCSP), for more accurate measure for the force needed to penetrate a target to a lethal depth while TCSA metric is more associated with cutting aspect. However, TCSP has a few disadvantages to directly apply to our case of stemmed points. The force and penetration depth are not only affected by the stone tip but also by the mass of the shaft, which cannot be known for archaeological stone-tipped weapons [@lombard2020tip]. -@sisk2011 also mentioned that TCSP cannot be applied to backed pieces that are hafted as projectile armatures.

\-@lee2019were first applied TCSA to stemmed points from Korea along with their use-wear analysis. The stemmed points for their research were discovered from Jingeuneul site located in southwest of Korea, where the largest number of stemmed points (n = 99) were found in a single site. For TCSA, they were only able to use ten stemmed points because they needed to select the stemmed points that retained the widest and thickest part of the specimens for TCSA metric and displayed diagnostic impact fractures for the use-wear analysis. The purpose of using TCSA in their research is to compare the values to North American dart tips and arrowheads. Their results show that the TCSA range of Jingeuneul stemmed points is relatively wide overlapped with both North American dart tips and arrowheads. Inspired by -@lee2019were in addition to other prior research that developed and applied TCSA into archaeological materials from other parts of the world, we explore stemmed points from all over South Korea to better understand their function to understand the role that stemmed points represent during the Late Paleolithic.

@@

To answer our research questions, What were stemmed points used for? How diverse were their functions? What are the temporal patterns in stemmed point functions? Are these temporal patterns associated with population dynamics or environmental processes? We calculated TCSA values for stemmed points and explored the usage for the stemmed points depending on raw materials, location, and size. We applied radiocarbon dates to the results to examine the temporal patterns. In the discussion, we compared those temporal patterns with demographic and environmental proxies. We used the results of summed probability distributions (SPD) and annual temperature changes from -@park2022did which covers our target period and area. Following the TCSA ranges for different weapon-delivery systems that -@lombard2021variation created by summarizing -@wadley2008segment' study of backed microliths (Table 1?), we excluded 12 artifacts from our dataset. Because the TCSA values of those 12 artifacts from Bupyeongri, Hwadaeri, Haga, Sinhwari, Gokcheon, and Sachang sites are greater than 257. Their TCSA range is from 263 to 2622. Therefore, a total of 161 stemmed points from 32 assemblages unearthed from 26 sites are used for the following analysis.

@@

The introduction of new types reflect major implications for social transformation. For example, the bow and arrow enabled individuals to work without hunting teams and provided opportunities for non-elite hunters to produce their own subsistence or pursue individual wealth [@bettinger2013effects; @angelbeck2014faustian; @rorabaugh2015timing]. In evolutionary terms, we assume that human groups carefully selected a specific stone tool based on its functional advantage over other alternatives once they acquired the necessary physiological and cognitive abilities to innovate [@lombard2022tip]. The choice of weapon-tip types, therefore, reflects the socio-environmental circumstances that they had to encounter and overcome. -@eren2022plains claim that Clovis points were more variable in shape than Folsom points because Clovis foragers were exposed to largely unfamiliar landscape and employed their points as multifunctional tools that perform a wider range of tasks including cutting and sawing. Folsom points show a narrow range of variation, indicating a specific task. They used TCSA to assess the morphological variance of the points as well as understand the potential functional costs or benefits of one artifact variant relative to another.

Similar to -@eren2022plains's approach, we hypothesize two scenarios that explain the regional or/and temporal pattern of TCSA range. If stemmed points represent a narrow TCSA range, people produced tools that performed certain tasks. There could be less surprise in the surrounding environment. On the other hand, a wide TCSA range indicates that stemmed points could be multifunctional tools and people might need to be prepared for unexpected situations such as moving into an unfamiliar landscape or their patch becoming less productive. We expect to observe various TCSA ranges, which reflect morphologically and functionally different stemmed points, across different regions or environments. However, we acknowledge that lithic morphology is highly dependent on the quality of raw materials and their availability.

@@

To gain a general idea of the role of stemmed points, we calculated TCSA for all stemmed points. @fig-tcsa-all-sp shows the mean TCSA is `r round(mean(tcsa$min_tcsa), digits=1)`, and standard deviation of TCSA is `r round(sd(tcsa$min_tcsa), digits=1)`. Overall we see a wide variation of TCSA values in this boxplot. According to -@lombard2021variation TCSA ranges (Table 1), Korean stemmed points are mostly in the categories of javelins and thrusting spear tips, with smaller numbers as dart tips and arrowheads. Among the weapon-tip types, only poison arrowheads appear to be absent from these Korean assemblages.

@@

In order to explore the relationship between raw material and the function of the stemmed points, we calculated TCSA per raw material type. @fig-tcsa-raw-materials shows that about half of the stemmed points were made from shale and its TCSA range is widely distributed. Acidic volcanic rocks tend to have skewed lower TCSA values. The other raw materials including hornfels, rhyolite, and tuff show a wide distribution, similar to shale. The category of ‘other’ raw materials include porphyry, trachyte, felsite, chert, quartz, quartzite, granite, mudstone, and unidentified rocks. Overall, there is no clear pattern of TCSA among raw materials.

>>>>

The stemmed point is the first composite tool in the Korean Paleolithic that requires two different parts, consisting of a stone point and shaft, presumably made out of wood [@seong2008tanged]. Using blades enables mass production for this composite tool and its shape becomes more standardized [@lee2015characteristics; @park2022did]. Therefore, @o2018origins claims that understanding the emergence of projectile technology gives insights to greater cultural, evolutionary, and behavioral cognitive flexibility.

Since the stemmed points represent the beginning of the Korean Late Paleolithic, investigating their origin(s) becomes a key for understanding the technological transition from the Early Paleoltihic, modern human dispersals, and existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern]. There has been a northworthy debate about explaining the origin(s) of stemmed points and it can be summarized into two competing models: 'heterogenic' migration [@bae2010origin], in situ evolution [@seong2009emergence]. The migration model claims that the new blade industry including stemmed points, and the earlier coarse flake tradition including large cores, polyhedrons, choppers, and even handaxes, came from different origins (e.g. 1: North route: Siberia, Mongolia, or other regions of northeast China, and 2: South route: southern China) as the result of continuing influx of modern human migration from two routes. The 'in situ' model argues that stemmed points and other Late Paleolithic technologies, including blade industries, autonomously emerged in the south of the Korean peninsula, with no apparent external influence. The difference between the two models comes from how to understand a few early sites that contain stemmed points but no blades and blade coares. The 'in situ' model claims that the stemmed points appeared before blades by acknowledging those early sites including Bonggok, Songamri, Yonghodong, and Hwadaeri sites while migration model supports the traditional definition of the Late or Upper Palaeolithic in Europe and Africa that started with emergence and establishment of blade technology [@seong2016eastern].

Stemmed points are important proxies to understand human behaviors during the Late Paleolithic. @park2022did examine mobility and site occupation patterns through lithic assemblages and concepts of human behavioral ecology and find that forager groups using stemmed points may have been associated with occupation of marginal or extreme environments compared to the groups with no stemmed points. Also, they were more associated with expedient technologies, indicating residential and less mobile behaviors. @Chong_2021 claims that the morphological variation of the stemmed points along with tool types in lithic assemblages, assemblage size, use of raw materials, and types of blanks could represent the characteristics of occupations such as "limited activity station" and "residential base camp." For example, stemmed points with high morphological variations in a tool size, shape of edge, degrees of damage, and types of edge retouching from Yongsandong site indicate that the site was used for specific or limited activities such as hunting [@kim2004yongsan; @seong2015diversity; @Chong_2021].

@@

Though the stemmed points are generally assumed as hunting armatures [@lee2019were; @lee2011study; @seong2008tanged; @lee2002new; @chang2013human], it is difficult to understand the function of the tools without knowing their complete shape attached to other components. Preserved wooden components of projectile tools are too few and rare to generalize the overall scale and variability of the tools used in prehistory [@shea2006origins]. @Lee_Kong_2002 claimed to name the stemmed points as 'stemmed tools' because of the diverse shape of stemmed points, which could represent different functions as well as the uncertainty of complete tool shape. Other researchers agree that some stemmed points that are not symmetrical and have retouch on one side of the artifacts or have denticulate blades on one side should not be called as stemmed points but stemmed knives, stemmed side-scrapers, stemmed end-scrapers, or stemmed burins [@seong2008tanged; @Lee_Jang_2011; @Kim] @fig-sp-hand-made: 7-8.

@@

The tip cross-sectional area (TCSA) has been used as a ballistically significant standard to discriminate different functional classes of projectile armatures such as spearthrower (a.k.a. atlatl) dart tips, arrowheads and large experimental stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins]. TCSA metric indicates the part of the tool that cuts the target's hide and relates to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation]. This method was first proposed by @hughes1998getting and and tested by @shea2006origins through comparing archeological examples to ethnographically collected samples of known use. @shea2009impact also applied this approach to compare projectile points from Africa, the Levant, and Europe and claimed that projectile weapons first appeared in Africa. The main advantage of the TCSA is, regardless of point type, only maximum width and thickness measurements are needed for calculating the TCSA value (0.5 × maximum width × maximum thickness) [@sisk2011african; @lombard2020tip]. Later, @sisk2011african proposed an alternate measure, tip cross-sectional perimeter (TCSP), for more accurate measure for the force needed to penetrate a target to a lethal depth while TCSA metric is more associated with cutting aspect. However, TCSP has a few disadvantages to directly apply to our case of stemmed points. The force and penetration depth are not only affected by the stone tip but also by the mass of the shaft, which cannot be known for archaeological stone-tipped weapons [@lombard2020tip]. @sisk2011 also mentioned that TCSP cannot be applied to backed pieces that are hafted as projectile armatures.

@lee2019were first applied TCSA to stemmed points from Korea along with their use-wear analysis. The stemmed points for their research were discovered from Jingeuneul site located in southwest of Korea, where the largest number of stemmed points (n = 99) were found in a single site. For TCSA, they were only able to use ten stemmed points because they needed to select the stemmed points that retained the widest and thickest part of the specimens for TCSA metric and displayed diagnostic impact fractures for the use-wear analysis. The purpose of using TCSA in their research is to compare the values to North American dart tips and arrowheads. Their results show that the TCSA range of Jingeuneul stemmed points is relatively wide overlapped with both North American dart tips and arrowheads. Inspired by @lee2019were in addition to other prior research that developed and applied TCSA into archaeological materials from other parts of the world, we explore stemmed points from all over South Korea to better understand their function to understand the role that stemmed points represent during the Late Paleolithic.

@@

To answer our research questions, What were stemmed points used for? How diverse were their functions? What are the temporal patterns in stemmed point functions? Are these temporal patterns associated with population dynamics or environmental processes? We calculated TCSA values for stemmed points and explored the usage for the stemmed points depending on raw materials, location, and size. We applied radiocarbon dates to the results to examine the temporal patterns. In the discussion, we compared those temporal patterns with demographic and environmental proxies. We used the results of summed probability distributions (SPD) and annual temperature changes from @park2022did which covers our target period and area. Following the TCSA ranges for different weapon-delivery systems that @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths (Table 1?), we excluded 12 artifacts from our dataset. Because the TCSA values of those 12 artifacts from Bupyeongri, Hwadaeri, Haga, Sinhwari, Gokcheon, and Sachang sites are greater than 257. Their TCSA range is from 263 to 2622. Therefore, a total of 161 stemmed points from 32 assemblages unearthed from 26 sites are used for the following analysis.

@@

The introduction of new types reflect major implications for social transformation. For example, the bow and arrow enabled individuals to work without hunting teams and provided opportunities for non-elite hunters to produce their own subsistence or pursue individual wealth [@bettinger2013effects; @angelbeck2014faustian; @rorabaugh2015timing]. In evolutionary terms, we assume that human groups carefully selected a specific stone tool based on its functional advantage over other alternatives once they acquired the necessary physiological and cognitive abilities to innovate [@lombard2022tip]. The choice of weapon-tip types, therefore, reflects the socio-environmental circumstances that they had to encounter and overcome. @eren2022plains claim that Clovis points were more variable in shape than Folsom points because Clovis foragers were exposed to largely unfamiliar landscape and employed their points as multifunctional tools that perform a wider range of tasks including cutting and sawing. Folsom points show a narrow range of variation, indicating a specific task. They used TCSA to assess the morphological variance of the points as well as understand the potential functional costs or benefits of one artifact variant relative to another.

Similar to @eren2022plains's approach, we hypothesize two scenarios that explain the regional or/and temporal pattern of TCSA range. If stemmed points represent a narrow TCSA range, people produced tools that performed certain tasks. There could be less surprise in the surrounding environment. On the other hand, a wide TCSA range indicates that stemmed points could be multifunctional tools and people might need to be prepared for unexpected situations such as moving into an unfamiliar landscape or their patch becoming less productive. We expect to observe various TCSA ranges, which reflect morphologically and functionally different stemmed points, across different regions or environments. However, we acknowledge that lithic morphology is highly dependent on the quality of raw materials and their availability.

@@

To gain a general idea of the role of stemmed points, we calculated TCSA for all stemmed points. @fig-tcsa-all-sp shows the mean TCSA is `r round(mean(tcsa$min_tcsa), digits=1)`, and standard deviation of TCSA is `r round(sd(tcsa$min_tcsa), digits=1)`. Overall we see a wide variation of TCSA values in this boxplot. According to @lombard2021variation TCSA ranges (Table 1), Korean stemmed points are mostly in the categories of javelins and thrusting spear tips, with smaller numbers as dart tips and arrowheads. Among the weapon-tip types, only poison arrowheads appear to be absent from these Korean assemblages.

@@

In order to explore the relationship between raw material and the function of the stemmed points, we calculated TCSA per raw material type. @fig-tcsa-raw-materials shows that about half of the stemmed points were made from shale and its TCSA range is widely distributed. Acidic volcanic rocks tend to have skewed lower TCSA values. The other raw materials including hornfels, rhyolite, and tuff show a wide distribution, similar to shale. The category of 'other' raw materials include porphyry, trachyte, felsite, chert, quartz, quartzite, granite, mudstone, and unidentified rocks. Overall, there is no clear pattern of TCSA among raw materials.
